---
title: "R Tutorial"
author: "Journey Keen"
date: "April 2026"
output:
  pdf_document:
    keep_tex: true
    pandoc_args: ["--listings"]
header-includes:
  - \setlength{\parindent}{15pt}
  - \setlength{\parskip}{0pt}
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{listings}
  - \usepackage{hyperref}
  - \hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}
  - \lstset{language=R,basicstyle=\ttfamily\small,numbers=left,numberstyle=\tiny,stepnumber=1,numbersep=6pt,breaklines=true,breakatwhitespace=false,showstringspaces=false,columns=fullflexible,keepspaces=true}
---

```{r setup, include=FALSE}
  knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    comment = "##"
  )
  
  library(data.table)
  options(width = 100)
```

\section{Introduction}
This was made so that students in the Minneapolis Fed's BEE Program have an introduction and some reference to data analysis with R, especially with graphing.

You, the student, should follow along. I will be placing the exact code and output for these notes in this document, but you should produce the code yourself to verify that you get the same output. Along with this pdf document, you should also download the files "example_file.csv" and "percentiles_combined.csv" for your scripts to work. You will also need the directory you put them in (which, for now, should be in the same folder as this pdf you downloaded and are reading, as well as your R script you start to follow-along).

Excel is nice because you can see the underling data right in front of you when you perform any analysis with the data. This is in contrast to a lot of computer programs where the data needs to be "printed" out to see anything. R allows a little bit of both, where your programming that reads like a left-indented text with a lot of newline characters and data are like tabs in an internet browser. You can switch between them pretty easily. What makes R (and other statistical programming languages) better is that you can do a lot more with them and the analysis is more ``systematic.''

What this means is that R is a programming language (kind of). You can create variables and data, and you interact with them via text like this (do \textit{not} run this code; it is for reference):

\hypertarget{intro-example}{}
```{r intro_example}
#### Do NOT type this -- Example only ####
example_data = data.table(year=2005:2007, value = 1:3)

x = 7
y = 4
z = x + y
```

Here the data "table" (which can be thought of as a dataset in excel, with rows and columns) and the variables \texttt{x}, \texttt{y}, and \texttt{z} are created. \texttt{z} is created from other variables. This is part of a "script," which you save and can run later to produce exactly the same data and, later, output. 

You will work in Rstudio, which is just a nice interface for using R. I attached a screenshot of what it looks like with my annotations for the four main sections in Rstudio. 

\includegraphics[width=\textwidth]{pictures/rstudio_ss_annotated.png}

When you have loaded-in data you can click the tabs between the programs and data like here, where the "R\_tutorial.R" is a program and the data is "data_example.'' It contains rows and columns, where the columns are names, just like in Excel. 

\includegraphics[width=\textwidth]{pictures/example_data.png}

The "programs" section of Rstudio is where the programs can be viewed and edited and where data can be viewed. Unlike in Excel, data cannot be changed by clicking on the data and typing anything: It must be changed by an actual program (or in the "console" section, but ignore that for now). The R text earlier was a quick example. 

The "environment" section shows which variables and tables you have created (and so are "in memory"). You can think of it as R keeping a log of the work you have done so you know what is loaded. In the annotated example it is empty. To view a data table you can just click on it and Rstudio will put it in as a tab in the programs section. You can also view data in the environment via the "console" section.

The console section is how you run code one line at a time, rather than part of a programming script in the programs section. This is useful when you might want intermediate code that is not to be saved to a program/script. For example, you might not want to save \texttt{x*y} to the environment or your larger project, but while coding you want to see what it looks like real quick. It is also where a lot of output goes, at least for our purposes here. In the earlier program we saved the variable \texttt{z=x+y} to our environment, but if we just wrote \texttt{x+y}, which does not save the value of x plus y to a variable but just computes it, then the output will be printed to the console rather than being saved to the environment as a variable, \texttt{z}. Here is what the console looks like if we created \texttt{x} and \texttt{y} in the console and then just wanted to quickly look at the value of \texttt{x+y}:

\includegraphics[width=\textwidth]{pictures/console_ss.png}

You can do the same by typing in each "line" (which are marked by the \texttt{>}) and pressing enter. Note that output has the \texttt{[1]} in front. Now, this is a silly example. We can do $7+4$ in our head pretty easily: Don't give-in and use ChatGPT or Claude for that, even if it is a homework assignment and it is getting late. This is just to demonstrate how it works. We can just as well put much more complicated things in the console that are intermediate checks for something that we do not want in our actual script. When writing the script to create some example graphs for this text I at one point typed this into the console to look at what data transformations I was creating:
```{r console_lookup, eval=FALSE}
#### Do NOT type this -- Example only ####
dt[year==1998 & group_var == "xall" & inc_var=="WS" & pctl==50, .(level, samp, geo_var, geo_var_val, group_var_val, pce, value, value_real)]
```

\noindent You do not need to know now what this means, but the point is that it it produces non-trivial output that was useful for me to see. Here is the actual output:

```{r console_lookup_output, echo=FALSE}
file_name = "C:/Users/IRJZK01/Desktop/IDDA/combined_data/percentiles_combined.csv"
dt = fread(file_name)
dt[year==1998 & group_var == "xall" & inc_var=="WS" & pctl==50, .(level, samp, geo_var, geo_var_val, group_var_val, pce, value, value_real)]
rm(dt)
```

The last section is what I called "Plots and Help." The console is where data output goes, and this is where plotting and graphical output goes. If you know a command but want the specifics of it, you can also type \texttt{?command} to get a file that explains the command. If you want to be more vague (e.g. if you forgot the actual name of the command you want to use) then you can use two question marks: \texttt{??command\_guess}. Yes, these helper things have mostly been replaced by AI. Sometimes they are quicker, though. You will rarely use the other tabs in this section and probably not at all in these two weeks.

\section{R Basics}
Here I will introduce how to actually program in R. You got a very small taste of it in the last section.

The first thing to do is to load-in common "packages." A package is just more of R you can use. Packages like downloading the Uber app on your phone, whereas base R has just the pre-installed apps, like messages or calls. You download apps and packages so your phone and R can do more. Today we will only use two packages, \texttt{data.table} and \texttt{ggplot2}. The first gives us access to Excel-like tools and the second to graphing and plotting tools. To install them, we use \texttt{install.packages()}, and then put the name of the package inside the parenthesis in quotes. This is like an app store: You only need to install them once. Do this:

```{r, eval=FALSE}
install.packages("data.table")
install.packages("ggplot2")

library(data.table)
```
\noindent The part \texttt{library()} is to actually load the package into your current R environment. Although you only need to install them once, you will need to put this at the top of every R script you have. We did not load-in \texttt{ggplot2} here because we will do it later when I introduce graphing.

\subsection{Variable Creation}
To create a variable you have a name for it and information to attach to it. I use "information" very loosely here; it can mean a real number, a vector, a matrix, or a table (variables and rows as in excel). When creating, you go by the format
\[ \texttt{variable\_name = information}\]
You can use the characters \texttt{<-} instead of the equals sign, \texttt{=}. They mean the same thing: Assign the information on the right to the name on the left. For example, \textcolor{blue}{\hyperlink{intro-example}{earlier}} \texttt{variable\_name} was "x" and \texttt{information} was "7." 

The three most common math operations are as normal, \texttt{+}, \texttt{-}, \texttt{*}, and \texttt{/}, for addition, subtraction, multiplication, and division, respectively.

\subsection{Logical Operators}
Next are logical operators. They check if something is true or false, which in R are the logical values \texttt{TRUE} (or just \texttt{T}) and \texttt{FALSE} (or just \texttt{F}). When working with data we frequently want to subset the data to something, where we only care about what we want to subset to. For instance, if you have data from $1990$ to $2025$ but only want to look at data in-and-around the pandemic, then you might to subset your data to years $2018$ to $2023$. If our data table has a variable \texttt{year}, then we need to get the rows where this variable is in this interval of years. That is a true or false condition. 

The common ones are
\begin{table}[!h]
    \centering
    \begin{tabular}{l|c|c|c|c|c|c}
       Operator  & \texttt{\&} & \texttt{|} & \texttt{==} & \texttt{!=} & \texttt{<}, \texttt{>} & \texttt{<=}, \texttt{>=} \\
       \midrule
       English Name  & And & Or & Equals & Not Equals & Less than, Greater than & Less than or equals...
    \end{tabular}
    %\caption{Caption}
    %\label{tab:placeholder}
\end{table}
For instance, \texttt{7==5} returns FALSE. \texttt{7==7} returns TRUE. We use the "and" and "or" operators to chain multiple expressions together, for example to check a condition for one variable and/or another condition for another variable. \texttt{TRUE == (TRUE | FALSE)} returns \texttt{TRUE}, since the right-hand-side evaluates to true as an "or" condition. But, \texttt{TRUE == (TRUE \& FALSE)} returns \texttt{FALSE}, since the right-hand-side then is false. An example with variables in a data table is \texttt{year <= 2017 \& color=="Blue"} which returns rows where the year variable is less than or equal to $2017$ \textit{and} the color variable is exclusively the text "Blue". If we instead had \texttt{year <= 2017 | color=="Blue"}, then we will get rows where the year variable is less than or equal to $2017$ \textit{or} where the color variable is the text "Blue". Rows then only need one of, or both, of these conditions. 

\subsection{Vectors}
Data science is just dealing with vectors and matrices. For our purposes a vector is just a "list" of another object, like real numbers, integers, characters, or even other vectors. In R we create vectors with \texttt{c()} like this:

```{r vectors}
vec_numbers = c(1,2,3,4)

vec_characters <- c("A", "B", "word")

vec_numbers_alt = c(5,6,9,10)

length(vec_numbers)
```

I also made an example of using \texttt{=} and \texttt{<-} for assignment. Again, they do the same thing. I returned the length of \texttt{vec\_numbers} to demonstrate that we control the length of a vector when we create it. Here is the same but for \texttt{vec\_characters}. 
```{r vectors_length}
length(vec_characters)
```
To access the information in a vector, you can pass an "index" to the vector. The character vector has "A" at the first index, "B" at the second, and "word" at the third. 
```{r vector_index}
vec_characters[3]
```
\noindent This returns "word", since that is at the third index. 

We can access the information in a vector in different ways than just the index of the information. In fact, using the exact index is rare. We typically want to access a vector where some condition is true and get the values there. We can pass a logical vector (so a vector filled with \texttt{TRUE}s and \texttt{FALSE}s) into a vector to get the information where the logical vector is true; that is, of the indices where the logical vector is true.
```{r vector_logical_index}
vec_characters[c(TRUE, FALSE, TRUE)]
```
\noindent returns a smaller vector, of length two, with the information of the character vector's first and third indices, since that is where the logical vector passed is true. Hence, you can see output is \texttt{c("A", "word")}. Try putting your own logical conditions in your own vectors. 

If you want a shortcut to create "sequential" vectors of the form $(n, n+1, \dots, n+j)$, then you can use \texttt{n:(n+j)}. This is the same as manually typing \texttt{c(n, n+1, } and then all the way to \texttt{, n+j-1, n+j)} for numbers $n$ and $j$. 

Vectors are important to be comfortable with because each column in a data table, our variables, are vectors. In fact, if you manually create a table (rather than loading-in an excel file or something), you are just telling R that a vector has a name, and then this named vector becomes a variable in the table. In our first example of R text
```{r example_data_again}
example_data = data.table(year=2005:2007, value=1:3)
```
\noindent we told R the variable \texttt{year} is the vector \texttt{2005:2007} and the variable \texttt{value} is the vector \texttt{1:3}. (Recall that these are just \texttt{c(2005,2006,2007)} and \texttt{c(1,2,3)}, respectively.) 

To give an example of where we would want logical conditions, create this in R:
```{r table_create}
table = data.table(
  gender=c("Male", "Female"),
  value =c(10, 13)
)
```
\noindent We might want to create a row from these two rows where our new row is a ratio of Female to Male values. For example, this data could be average earnings. The code to do this is below. Focus only on the \texttt{ratio = } part. The rest I will cover later. 
```{r table_ratio}
table[, .(
  ratio = value[gender=="Female"]/value[gender=="Male"]
)]
```
\noindent We access the row where gender is "Female" in the vector \texttt{value} and divide it by where this vector is "Male". Yes, you can just look at it and give the answer (or give your favorite AI software the male and female value and ask it what the ratio is, if you must) easily here. If you wanted to do this directly with vectors instead of the table format you could have
```{r vector_ratio}
gender = c("Male", "Female")
value = c(10, 13)
value[gender=="Female"]/value[gender=="Male"]
```
\noindent which returns $1.3=13/10$. 

In a more realistic but more detailed version this data would exist over many years and there could potentially be information not related to gender in the table. You would not want to scroll over it every year to get the two values you want and then divide them all, manually. This is why we use R.

\subsection{Organization}
One of the benefits to using statistical software like R is all of these saved scripts that are the blueprints of how to take some data and create other data, be it a regression/model summary, plots, or manipulated data to be used by another file. You will end-up creating a lot of files. You need to be organized in each file, across the files in a project you are working on, and across the projects you may or may not be working on. 

There is a lot of literature on how to stay organized here. It is extremely important, and in more recent years economists with computer science backgrounds (software engineers deal with programs with several thousand times the complexity in terms of how messy their files can potentially be compared to economists, and so they typically are quite good at this) have pushed for doing this better.

At some point you will have to do this. I do not think you should worry about it now, if this is your first time using programming or statistical software. The best way to learn this is probably the hard way: Burn your finger, say mean things, and figure out fire is hot. Poorly label a bunch of files, put dates onto files, do not comment on what your code is doing. Erase your work from your mind, come back a week later, and pick-up where you left off. Your finger is now burnt. So, waste an hour (with mean words) trying to find where and what you were working on. You will re-organize a bunch of code in the process of doing this. You will then break the code because you forgot what was supposed to be ran first and other things that part of the code requires to run. And then you will learn fire is hot and how to use a rag when grabbing a hot pan: You will learn how to organize code, your projects, and how to record your work and edits thereof. But not now. For now just focus on writing those first few basic scripts.

\subsection{General}

```{r, echo=F}
"
Not sure what else to include here. Probably need something about libraries/packages. Reading and writing data via csvs (which is all I will cover) is below. I could add a subsection on general formatting and standard syntax usage (i.e. my own). 

Edit: I should cover folder structure and directories.

Edit: I should cover basic string manipulation (spamming \texttt{paste0}). No \texttt{sprintf} in this tutorial.
"
```



\section{Data Tables}
You have already seen examples of collecting vectors, naming them, and putting them in a "table" in R. Once you view it, by clicking on it in the environment section, it looks a lot like an Excel sheet. Here you will see the huge improvements to working with R. You can edit existing variables, create new ones, and in general manipulate your data table much easier in R than Excel. The degree to which this is easier scales with the difficulty of the problem you are facing. 

\subsection{Reading and Writing Data}
You will almost never manually create a data table like how I did earlier, in order to create examples. The only time I ever do is to make small data with really weird cases to check my programming logic is correct (and to make these notes). You will typically be reading and writing to data stored on your computer, or, in more advanced cases, directly through the internet. Here I will only cover reading and writing to a "csv". It is basically an Excel file. When you load a csv on your computer it takes you to Excel. "csv" stands for "comma separated values". It just stores data as it is displayed in an Excel sheet, just the data in each cell is separated by a comma. In economics a lot of publicly available data is stored this way. It is probably the easiest and most universal way to store basic data, hence why we will be using it here. As your computing needs grow you will need to graduate from using csvs. But not today.

Say you have the csv in your computer with the file name 
\[ 
"C:/Users/person/Desktop/folder\_1/folder\_2/data\_folder/file.csv"
\]
The ".csv" at the end tells us it is a csv. We will load-in ("read") files of this type with \texttt{fread}, like
```{r fread_example, eval=FALSE}
dir = "C:/Users/person/Desktop/folder_1/folder_2/data/"
dt = fread(paste0(dir, "example_file.csv"))
```
\noindent You need to put your actual directory you stored the data from GitHub here, not my fake example. We now have the data table \texttt{dt} in our environment. Say we did a bunch of manipulation to \texttt{dt} and end-up with the table \texttt{dt\_manip}. We would want to save it ("write" it) to the computer in the same directory, \texttt{dir}. We will use \texttt{fwrite}:
```{r fwrite_example, eval=FALSE}
## Trivial manipulation. You do not need to understand what it means:
dt_manip = dt[group=="Female", .(category, average)]


output_file = "example_file_manipulated.csv"
fwrite(dt_manip, paste0(dir, output_file))
```

```{r dt_setup, include=FALSE}
## Remark: For the kiddos I will make this a file they download, so they do not see how I created it, here.
dt = data.table(
  category = c(rep("Gender", 6L), rep("Race", 9L), "Age"),
  group = c(rep(c("Male", "Female"), 3L), rep(c("White", "Black", "Hispanic"), 3L), "23"),
  average = c(10,13,  11,13,  12,14,
              14,14,14,  15,14,13,  14,16,17,
              10),
  year = c(2010,2010, 2011,2011, 2012,2012,
           2010,2010,2010, 2011,2011,2011, 2012,2012,2012, 2010)
)
dt[, gender := ifelse(category == "Gender", group, NA_character_)]
```

(If you are outputting into a different directory, say \texttt{dir\_manip}, then use that above instead of \texttt{dir}. You will usually want to do this: Separate inputs and outputs. See the organization section above.) 

\subsection{Viewing Data}
In this section \texttt{dt} will denote an arbitrary data table. Previously we have used names like \texttt{example\_data} and \texttt{table}. 

If you want to put the entire table in your programs tabs to view like an Excel sheet, then click the table in the environment. If your table is more than a few hundred rows or more than about twenty columns this will quickly become cumbersome, as you might, say, want to quickly look at row $100,000$ or the first fifteen rows where the variable \texttt{age} is $23$. This might not be in your table until row $100,001$, so you will not want to scroll down that far. Hence, we need to learn how to quickly view the part of the data we want. We will be using the scary console.

To print all rows, use \texttt{dt[]}. If there are a lot of rows this will print the first and last five of the table. To get specific row numbers (or just a number), put the number(s) of the rows you want inside the brackets. This works like accessing the information in a vector. 
```{r dt_view_rows, eval=FALSE}
dt[1]
dt[c(1,2,3)]
dt[1:3]

dt[17]
```
\noindent Line $1$ returns the first row, since the row indexing starts at one. Lines $2$ and $3$ do the same thing by returning the first three rows, since \texttt{c(1,2,3} and \texttt{1:3} are the same vector (of indices here). Line $5$ will return a table with missing (NA) values in the columns since there is no row $17$, since there are only $16$ rows. Here are the actual four outputs:
```{r dt_view_rows_output, echo=FALSE}
dt[1]
dt[c(1,2,3)]
dt[1:3]

dt[17]
```

Say we want the rows where the variable \texttt{group} is $23$, similar to the prior example. Then
```{r dt_view_group, eval=FALSE}
dt[group == "23"]
```
\noindent returns the rows where this is true. It is the same as passing a logical vector of true and false values into our table, where there are \texttt{FALSE}s where \texttt{group} is \textit{not} "23" and are \texttt{TRUE}s where \texttt{group} \textit{is} "23". I hope you see the point of the overly long sections on vectors and logical conditions now. 

Something similar holds with the columns. The columns are ordered like the rows are, with the leftmost column being $1$. Hence, you can print/return all rows of the second column (or the first and last $5$ rows of it, if the data is large) similarly, like:
```{r dt_second_col}
dt[, 2]
```
\noindent Note the comma: Rows are before the comma and columns are after, just like how we organize a matrix (do not worry if you do not know what that is). You can combine these to get even more specific. The following prints/returns the third row of the second column:
```{r dt_thirdrow_secondcol}
dt[3, 2]
```


You will typically only print specific rows of the table to see the first row, to remind yourself of what the table looks like (yes, you will forget at times). Usually you will want to pass logical conditions to view specific rows you are concerned about when coding, like the \texttt{group=="16"} above. Likewise, you will probably never use the column indexing again. In fact, you should not. This might be my first time doing that since learning R; it was to show you that the data table is just a bunch of vectors with some fancy code in the background holding it together. 

You will instead access the columns/variables by their names. Our second column is \texttt{group} (this is the main reason to print the first row with \texttt{dt[1]}, to remind yourself what the column names are), and so instead of putting a $2$ there we can put its name, like so:
```{r dt_group_column}
dt[group == "23", group]
```

What if you want multiple columns to be returned? The columns are vectors, so you can do this to return the columns \texttt{group} and \texttt{average}:
```{r dt_two_columns}
dt[group == "23", c("group", "average")]
```
\noindent This is annoying and slow to type with the quotes around them. Thankfully, we can do this:
```{r dt_two_columns_dot}
dt[group == "23", .(group, average)]
```
\noindent Note the dot in front. There is a technical name for doing that. Let the nerds worry about that. You still needs the parenthesis around the list of comma-separated variables names.

\subsection{Manipulating Data}
Let's say you want a column full of $1$s. I'm not sure why, but let's say you do. We add columns to the table with the \texttt{:=} operator (you will see this eventually in math classes; economists like to use $\equiv$; yuck). Assignment works like assigning information to a name at the beginning of this tutorial, but since this is a column it needs to go after the comma:
```{r dt_manipulate}
dt[, column_of_ones := 1]

dt[, column_of_ones_long := c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)]

dt[group=="23", one := 1]

dt[category=="Gender",
   F_M_ratio := average[gender=="Female"]/average[gender=="Male"],
   by = year]
```
\noindent Line $1$ is how you should do it. Line $3$ shows what R is really doing, since creating a new variable is the same thing as creating a new vector and putting it in the table. Note that the vector I assigned was length $16$, the length of the table. R will give us ("throw") an error if we attempt to assign a variable of another length.

Line $5$ is a sneak peak into how we combine the row and column operations other than just to view the a subset of the table. Like when we viewed the part where the condition \texttt{group=="16"} held, this first subsets to those rows, and \textit{then} we assign the variable there. The variable must exist on the other rows, but we did not assign those other rows any information. Hence, R assigns other rows (where \texttt{group != "16"} is true) missing values, NAs (to be precise, \texttt{NA\_character\_} values). 

Line $7$ combines what line $5$ did with an example of a more complicated manipulation. On the set where Male/Female makes sense (where \texttt{category=="Gender"}), we assign the F/M ratio for each year. We did nearly the same thing in a prior example. To do it for each year we needed the \texttt{, year} at the end, since otherwise the \texttt{average[gender==]} call would result in a vector, the values across all of the years. I will cover that in the next subsection. 

We can perform all other standard operations between variables (again, to be thought of as vectors) in our table. Here I will make a different table to demonstrate the more basic operations:
```{r DT_basic_ops}
DT = data.table(
  x=c(1,2,3,7,8),
  y=c(2,2,2,3,4)
)

DT[, z := x+y]
DT[, w := x-y]
# The ':='( just means everything inside the parentheses gets created, and we no longer need the colon, just the = sign. It can save space in your file.
DT[, ':='(u = x*y, v = x/y)]
DT[, log_y := log(y)]

DT
```
\noindent Once we create the table \texttt{DT} we create other variables with common operations. \texttt{z, w, u}, and \texttt{v} are the sum of, difference of product of, and quotient of \texttt{x} and \texttt{y} as vectors, respectively. I say "as vectors" because these are performed entry-wise: Each row has \texttt{z} as the sum of that row's values of \texttt{x} and \texttt{y}, and the same goes for the other three variables.\texttt{log\_y} looks a bit different, but it is conceptually the same. It is the natural logarithm function applied to each row's value of \texttt{y}.

The last two variables created are a bit special, since they are not computed entry-wise. The mean of \texttt{x} and variance of \texttt{z} are computed over the whole vector/variable, not just that row (as \texttt{x+y} is, for example). It is the same as computing the mean and variance over each variable as a vector, saving these values, and assigning every row that single value as we did with assigning ones before. You should check that this does the same:
```{r DT_summary_cols}
DT[, mean_x := mean(x)]
DT[, var_z := var(z)]

mean_x_value = mean(DT[, x])
var_z_value = var(DT[, z])

DT[, ':='(mean_x=mean_x_value, var_z=var_z_value)]
```
\noindent I do not print \texttt{DT} so you can check it.

\subsection{Grouping Data}
Earlier we had a mysterious \texttt{, by = year} at the end of our \texttt{dt} call:
```{r dt_FM_ratio_again, eval=F}
dt[category=="Gender",
   F_M_ratio := average[gender=="Female"]/average[gender=="Male"],
   by = year]
```
\noindent This "grouped" our data together for each year, so that when we accessed values for a specific gender it was comparing values in a single year (which is what made the operation possible in the first place). 

We do this all the time in data science: Group rows together by the year, or the month, or the gender, or the location, etc. and create a new variable that is defined over each possible value you can group by. The possible values for \texttt{year} were $2010, 2011$, and $2012$, so it created \texttt{F\_M\_ratio} over each of these years. We do this particularly often when we want summary stats for each group: Get the average of \texttt{x} for each month; the average of \texttt{y} for each county; the variance of \texttt{z} for each possible \textit{combination} of year, month, and county in our dataset. This last example notes that we can group by multiple variables at once. 
```{r DT_grouping_year}
DT = data.table(
  year=c(2024,2024,2024,2024, 2025,2025,2025,2025),
  month=c(1,1,12,12,  1,1,12,12),
  county=c("Hennepin","Ramsey","Hennepin","Ramsey",
           "Hennepin","Ramsey","Hennepin","Scott"),

  x=1:8
)

DT[, mean_year_x := mean(x), by = year]
```

We took the mean of \texttt{x} just grouping by the year. That is, we did not consider that the month or county could vary within the year. But, we can with
```{r DT_grouping_multi}
DT[, mean_year_month_x := mean(x), by = .(year, month)]
DT[, mean_year_month_county_x := mean(x), by = .(year, month, county)]
```
\noindent When grouping by multiple variables we list them like we did in Section $3.1$, with the dot and parenthesis (likewise, we can make it a vector of strings of the variable names, but this was here is quicker). Line $1$ groups by both year and month. Hence, we will take the average across every county in every possible \texttt{year}$\times$\texttt{month} combination. Line $2$ groups by year, month, and county. Note that in this example there is only ever one value of \texttt{year}$\times$\texttt{month}$\times$\texttt{county}: There is never more than one row that has the same year, month, and county. Hence, the average will just be the value of \texttt{x} itself, as you should verify.

\subsection{Collapsing (Summarizing) Data}
Everything you have seen thus far involves adding a new column to the table, with the \texttt{:=} operator (for potential comp. sci. people this really means the columns are added by reference). Doing so does not change anything already in the table. But sometimes we might not want that. For instance, in our last example, if we just wanted the annual mean of \texttt{x}, then using \texttt{:=} adds a column that could be redundant if we do not need this extra column for any other purpose. You will encounter this a lot, where you want some summary statistic(s), but it does not need to go in your main data. (In fact, cluttering your main data table with summary statistics that you do not need for future manipulation but only to view is actively discouraged.) 

To do so in the simple version we just drop mention of the \texttt{:=}. 
```{r DT_collapse}
DT[, mean(x), by = year]

DT[, .(mean_year_x = mean(x)), by = year]
DT[, .(mean_x_year_county = mean(x)), by = .(year, county)]
```
\noindent Line $1$ accomplishes what line $10$ did two examples ago in terms of the values it produces. You should check that they are the same. But, you will notice that there are now only two rows, the number of \texttt{year}s in the table. We did not use \texttt{:=}, so there was not a new column added, so any information (rows and columns) that are not the data we are manipulating to create summary statistics (\texttt{x}) and the variable we are grouping by and its levels (\texttt{year}) get dropped. In line $4$ here I group by year and county, so summary table will now have three columns: \texttt{year} and \texttt{county} that has every possible combination of them in the table, and \texttt{mean\_x\_year\_county} that is the mean of \texttt{x} over the rows that match each combination. 

In line $3$ I also introduced how you can name the variable you are summarizing. If you do not supply a name it will default to things like \texttt{V1}. Note that it is still inside the dot and parenthesis. This also applies to when we want to summarize multiple variables or in multiple ways over the same grouping:
```{r DT_multiple_summaries}
# Add a variable y so we have multiple numeric variables
DT[, y := 5:12]

DT[, .(mean_year_x = mean(x), mean_year_y = mean(y)), by = year]
DT[, .(mean_year_x = mean(x), var_year_x = var(x)), by = year]
```


Optional: If a variable is constant within each combination of values of what you are grouping by, then you can include it in the summary stats by including it in the variables you are creating. For instance, if \texttt{code} is the same in every year, then you can write
```{r DT_code_constant, eval=FALSE}
DT[, .(mean_year_x = mean(x), code), by = year]
```
\noindent and still have \texttt{code} appear as a variable. Now, if \texttt{code} varies within any level of \texttt{year} (e.g. in $2024$ it has two possible values, $1$ and $2$), then our data will no longer collapse down into unique years of the table, as we have not supplied a way to summarize it. 

\section{Examples and Practice}
Here we will load-in data that you will use as part of one of the BEE lectures. It has income percentiles for different demographic groups (age groups, race, foreign- or U.S.-born, and gender, as well as product variables thereof) in the United States for different income variables, in different states, and for some different underlying population samples. 

We load up the data and look at the first row to see the variables and the type of data:
```{r pctls_read_firstrow}
file_name = "C:/Users/IRJZK01/Desktop/IDDA/combined_data/percentiles_combined.csv"
pctls = fread(file_name)

pctls[1]
```
\noindent (In your output, if your console section is sufficiently wide, then this should output on one line. Here it looks more messy than it normally is.)

```{r commented_out, include=FALSE}
if (F) {
  #\begin{table}[!h]
  #    \centering
  #    \begin{tabular}{ccccccccccccc}
  #        year & level & samp & inc\_var & geo\_var & geo\_abb & group\_var & group\_var\_val & pctl & value & pce & value\_real \\
#  
  #        $<$int$>$ & $<$char$>$ & $<$char$>$ & $<$char$>$ & $<$char$>$ & $<$char$>$ & $<$char$>$ & $<$char$>$ & $<$num$>$ & $<$num$>$ & $<$num$>$ \\ 
  #        1998 & mafid & all\_1040\_mafid & WS & usst & 0 & US & xaged & 16to24 & 10 & 6969 & 0.6856143 & 10164.61
  #    \end{tabular}
  #    %\caption{Caption}
  #    %\label{tab:placeholder}
  #\end{table}
}
```


For our output we care about the "W-2 prime age working" sample, which corresponds to a \texttt{samp} of "prime\_age\_working". The income variable, \texttt{inc\_var} is "TC", and we only want U.S.-level data, so we want \texttt{geo\_var} to be "usst". We also only want percentiles in $\{25,50,75,90,95,98,99,99.9\}$, so we can filter to that interval (first you should check that we can, indeed, take it as an interval). We are comparing Black male and Black female percentiles to White male percentiles, so we also want to filter to \texttt{group\_var\_val}s of "NH\_Black\_Male", "NH\_Black\_Female", and "NH\_White\_Male". By comparison we mean take the ratio of the Black percentiles to the White ones, so for each percentile in each year we want to take the ratio of the two. Hence, we will be grouping by \texttt{year} and \texttt{pctl}. 

I subset the entire data \texttt{pctls} to smaller data, \texttt{data}, so then we only have the rows we want. I then split the data into two again, the male-male comparison and the female-male comparison. Here is the code:
```{r pctls_filter_ratios, eval=TRUE}
data = pctls[samp == "prime_age_working_w2" & inc_var == "TC" & group_var=="xredXsex" &
               group_var_val%in%c("NH_White_Male","NH_Black_Male","NH_Black_Female") &
               geo_var == "usst" & 25 <= pctl & pctl <= 99.9
]
data_male = data[group_var_val%in%c("NH_White_Male","NH_Black_Male"),
                 .(ratio = value[group_var_val=="NH_Black_Male"]/value[group_var_val=="NH_White_Male"]),
                 by = .(year, pctl)]
data_female = data[group_var_val%in%c("NH_White_Male","NH_Black_Female"),
                   .(ratio = value[group_var_val=="NH_Black_Female"]/value[group_var_val=="NH_White_Male"]),
                   by = .(year, pctl)]
```

Graphing for most of our basic purposes in two weeks will be easy. It is also easy to try to spend too much time making plots look trivially nicer. Try not to do that for what are rough drafts. Here is the code for our first graph:

```{r male_plot}
library(ggplot2)

data_male[, pctl := as.factor(pctl)]
ggplot(data_male, aes(x=year, y=ratio, color=pctl, group=pctl)) +
  geom_point() + geom_line() +
  theme_minimal()
```

And here is the second.
```{r female_plot}
data_female[, pctl := as.factor(pctl)]
ggplot(data_female, aes(x=year, y=ratio, color=pctl)) +
  geom_point() + geom_line() +
  theme_minimal()
```

We needed to put the variable \texttt{pctl} as a "factor" for reasons you should not worry about now. But, any time you have "groups" of something (so to create this you grouped by it, and in the plotting you made the colors group by it), you should do this before plotting. The general syntax is
```{r plotting, eval=FALSE}
ggplot(dt, aex(x = x_variable, y = y_variable, colour = group_variable, group = group_variable)) +
  geom_point() + geom_line() +
  theme_minimal()
```
\noindent Our x variable was \texttt{year}, our y variable was \texttt{ratio}, and we wanted our data to connect the dots to form a line based upon \texttt{pctl}, so we told R to color by it and group them together.

\[\textrm{... Examples to be added ...}\]

